Non-word repetition in Yélî Dnye

### 2020-12-23 manuscript.Rmd
midway through results, knits

Third, there are ongoing discussions as to what the key factors structuring individual variation are. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review on the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

<!-- a2m do you remember: why did we remove bilingual exposure? -->

@@

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early]. While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day). It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segments of each school year. Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

@@

- The same three items were selected as practice items and used in all 40 elicitation sets.
- Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sublists). Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 vs. 23).
- Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1-, 2-, and 4-syllable), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.

@@

The second phase of the experiment involved running the child through the list of test items randomly assigned to them. This was done in the same manner as the practice items: the stimulus was played over the headphones, and then the child repeated it aloud. NWR studies vary in whether children are allowed to hear and/or repeat the item more than one time. We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make a further attempts if their first attempt was judged erroneous in some way by the assistant. The procedure worked as follows: When the child made an attempt, the assistant indicated to the experimenter whether the child's production was correct or not. If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam recorder and the group moved on to the next item. If not, the child was allowed to try again, with up to five attempts allowed before moving on to the next item. Children were not asked to make repetitions if they did not produce a first attempt. In total, test sessions took approximately XX minutes on average, with XX of those minutes attributed to practice and XX to the actual test list.
<!--M2A: Alex, could you please estimate the above XX cases and fill in?-->

@@

The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat],
<!--M2A: Alex, I added a bib entry for Praat but it's the most recent version. Could you please correct it to the version (approximately if not precisely) that you used?-->
then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files. The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions. The assistant indicated in a notebook whether each production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English. The assistant also provided some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.

@@

In addition to a binary word-level exact repetition score we scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer [the target or the attempt; see also @cristia2020infant]. Although previous work does not use distance metrics, we additionally report these. For instance, recall our example of an English target being /bilik/ with an imagined response [bilig]. We would score this response as follows: at the whole item level this production would receive a score of zero (because the repetition is not exact); at the phoneme level this production would receive a score of 80% (4 out of 5 phonemes repeated exactly); and the phone-based Levenshtein distance for this production is 20% (because 20% of phonemes were substituted or deleted).  
<!--M2A: A concrete scoring example would be a good idea here-->
<!-- a2m added! actually, I haven't changed anything but I've been hesitant about using levenshtein's distance because it's the complement of phoneme based scoring -- so shall we drop it?  -->
<!-- M2A: three thoughts on this, which you can take as you like:
1) our audience might make this determination. segment-match score (phoneme-level accuracy) is probably more straightforward and might therefore be preferred by psychies
2) LD has the advantage of making D/I/S relevant as a next step in the analysis
3) we should definitely choose one, I agree, and report the other in supp mat tables-->

<!--M2A: So what's the final determination here, keeping in mind we're now aiming for LDR?-->

Additionally, we estimated the typological frequency of all phonological segments used in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible]. For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory. While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at time of writing, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt. In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication; see also [web.phonetik.uni-frankfurt.de/upsid_info.html](http://web.phonetik.uni-frankfurt.de/upsid_info.html)). It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database. Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat these unattested phones as having a frequency of 1, with a (rounded) percentage of 0%.

Finally, when describing children's patterns of errors, we take all repetitions of a given target into account. We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
<!-- a2m in the results, I put analyses based on all as well as just on first attempt. If we do just the latter, then come back to change this para -->

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years). Among these children were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth.

[^1]: Years of education is difficult to precisely pinpoint in this context. Schooling beyond grade 8 takes place off-island and enrollment is tied to an ability to pay that year's school fees, seasonal interruptions to school opening times, familial obligations, and more. We asked parents to report their highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions. This method results in many esimates around typical "graduation" times (e.g., graduating from primary education before starting secondary education).

@@

We first checked whether whole-item NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions. We excluded `r excl` child who did not have data for one of these two types. As shown in Figure \@ref(fig:Fig1-first-vs-others), participants' mean word-level scores became more heterogeneous in subsequent repetitions. Surprisingly, whole-item NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p `r pval.display(round(myt$p.value,2))`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`). Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions.

@@

Taking into account only the first attempts, we averaged attempts by each of the `r dim(overall)[1]` children who had data for first attempts; their ages ranged from `r min(overall$Age)` to `r max(overall$Age)` (M = `r round(mean(overall$Age),1)`, SD = `r round(sd(overall$Age),1)`). The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`. Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%), Cohen's d = `r cohensd(mean(overall$Phoneme),sd(overall$Phoneme))`. The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted (the normalized Levenshtein distance is the complement of phoneme-based scores).

@@

```{r diffic,echo=F}
nasalVs=c(":a", ":â" ,":ê", ":i" ,":o")
oralVs=c("a", "â","e",  "é" ,"ê", "i","î" ,"o","ó","u")
simplexCs=c("m","ń","l","w","y","v","d","g","p","t","k","f","h","ch")
complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")
simplexCs.ipa=c("m","n","l","w","j","w","ṭ","g","p","t","k","f","h","tʃ")
complexCs.ipa=c("ṭp","tp","kp","km","kṇ","mp","ɣ","lβʲ")

@@

read.table("substitution_bank_try1.txt",header=T)->sub_bank
colnames(sub_bank)<-c("target","realization")
sub_bank$target=factor(sub_bank$target,levels=allVC)
sub_bank$realization=factor(sub_bank$realization,levels=allVC)

@@

read.table("deletion_bank_try1.txt",header=T)->del_bank
colnames(del_bank)<-c("target")
del_bank$target=factor(del_bank$target,levels=allVC)
del_tab= table(del_bank)[allVC]

read.table("match_bank_try1.txt",header=T)->match_bank
colnames(match_bank)<-c("target")
match_bank$target=factor(match_bank$target,levels=allVC)
match_tab= table(match_bank)[allVC]

@@

<!--M2A: coming back to these three comments below after I manage another full read-through (after I check demo) -->

<!-- a2m I think it's weird to put effect sizes here, right? technically I could do something like cohen's d based on mean and sd of the percentage of tokens... -->

<!-- **a2m regarding the use first attempts rather than the whole data set, I'm having second thoughts. To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers & it is more consistent. BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros... CF Option 3 further down** -->

<!-- **a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?** -->

<!--
M2A: I prefer a modification of option 2 and so am commenting this out:

**Option 1: just numbers, first attempts only **
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted, and `r sum(del_tab[c(simplexCs,complexCs)])` consonants.

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality). As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])`) than vice versa (`r sum(sub_tab[simplexCs,complexCs])`). Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones).

**Option 2: numbers and proportions, first attempts only**
-->

Next, we investigated patterns of deletion and substitution. Deletions were relatively rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets). Substitutions were somewhat more frequent, but still relatively rare. Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets)). Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (`r sum(sub_tab[oralVs,oralVs])` oral vowels (~`r mean(sub_pc[oralVs,oralVs],na.rm=T)`% of targets and `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels (~`r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% of targets). Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets. Interestingly, simple consonants were sometimes also substituted by complex ones; this happened `r sum(sub_tab[simplexCs,complexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all simple consonant targets. Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets). Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms (see below).

```{r diffic_all_attempts,echo=F}

read.table("substitution_bank.txt",header=T)->sub_bank
colnames(sub_bank)<-c("target","realization")
sub_bank$target=factor(sub_bank$target,levels=allVC)
sub_bank$realization=factor(sub_bank$realization,levels=allVC)

read.table("deletion_bank.txt",header=T)->del_bank
colnames(del_bank)<-c("target")
del_bank$target=factor(del_bank$target,levels=allVC)
del_tab= table(del_bank)[allVC]

read.table("match_bank.txt",header=T)->match_bank
colnames(match_bank)<-c("target")
match_bank$target=factor(match_bank$target,levels=allVC)
match_tab= table(match_bank)[allVC]

#this means: divide each column by corresponding number in denominator vector
#sub_pc has the same shape as the original sub_tab, namely targets as rows, realization as columns
sub_pc=round(sub_tab/c(denom_vec)*100,2)

del_pc=round(del_tab/denom_vec*100,2)
mat_pc=round(match_tab/denom_vec*100,2)

<!--
M2A: I prefer a modification of option 2 and so am commenting this out:

**Option 3: numbers and proportions, ALL attempts**
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted (about `r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` consonants deleted (about `r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels or about `r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% out of all oral vowel targets, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels or about `r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% out of all nasal vowel targets). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality or about `r mean(sub_pc[oralVs,oralVs],na.rm=T)`% out of all oral vowel targets; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality or about `r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% out of all nasal vowel targets). As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])` times or  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all complex consonant targets) than vice versa (`r sum(sub_tab[simplexCs,complexCs])` times  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all simple consonant targets). Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones or about `r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% out of all simple consonant targets, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones or about `r mean(sub_pc[complexCs,complexCs],na.rm=T)`% out of all complex consonant targets).

-->

>>>>

Third, there are ongoing discussions as to what the key factors structuring individual variation are. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review on the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

@@

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early]. While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day). It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segments of each school year. Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing]. Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

@@

- The same three items were selected as practice items and used in all 40 elicitation sets.
- Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sublists). Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 vs. 23).
- Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.

@@

The second phase of the experiment involved running the child through the list of test items randomly assigned to them. This was done in the same manner as the practice items: the stimulus was played over the headphones, and then the child repeated it aloud. NWR studies vary in whether children are allowed to hear and/or repeat the item more than one time. We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make further attempts if their first attempt was judged erroneous in some way by the assistant. The procedure worked as follows: When the child made an attempt, the assistant indicated to the experimenter whether the child's production was correct or not. If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam recorder and the group moved on to the next item. If not, the child was allowed to try again, with up to five attempts allowed before moving on to the next item. Children were not asked to make repetitions if they did not produce a first attempt. In total, test sessions took approximately six minutes, with the first minute attributed to practice and five minutes to the actual test list.

@@

The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat], then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files. The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions. The assistant indicated in a notebook whether each production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English. The assistant also provided some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.

@@

Previous work typically reports two scores: a binary word-level exact repetition score, and a phoneme-level score, defined as the number of phonemes that can be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer [the target or the attempt; as in @cristia2020infant]. Previous work does not use distance metrics, but we report these rather than the phoneme-level scores because they are more informative. To illustrate these scores, recall our example of an English target being /bilik/ with an imagined response [bilig]. We would score this response as follows: at the whole item level this production would receive a score of zero (because the repetition is not exact); at the phoneme level this production would receive a score of 80% (4 out of 5 phonemes repeated exactly); and the phone-based Levenshtein distance for this production is 20% (because 20% of phonemes were substituted or deleted). Notice that the phone-based Levenshtein distance is the complement of the phoneme-level NWR score. An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (additions were not attested).

Additionally, we estimated the typological frequency of all phonological segments used in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible]. For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory. While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors). Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication). Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0% (i.e., its cross-linguistic percentile is zero).

Finally, when describing children's patterns of errors, we take all repetitions of a given target into account. We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, insertions, and substitutions.

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years).

Included children's ages ranged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years). We included `r sum(inc$Sex=="B")` boys and `r sum(inc$Sex=="G")` girls.  There were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children. These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.

[^1]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.

@@

We first checked whether whole-item NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions. We excluded `r excl` child who did not have data for one of these two types. As shown in Figure \@ref(fig:Fig1-first-vs-others), participants' mean word-level scores became more heterogeneous in subsequent repetitions. Surprisingly, whole-item NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p `r pval.display(round(myt$p.value,2))`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`). Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions, with the exception of qualitative analyses of substitutions.

@@

Taking into account only the first attempts, we derived overall averages across all items. The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`. The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.

@@

```{r diffic-all-items,echo=F}
nasalVs=c(":a", ":â" ,":ê", ":i" ,":o")
oralVs=c("a", "â","e",  "é" ,"ê", "i","î" ,"o","ó","u")
simplexCs=c("m","ń","l","w","y","v","d","g","p","t","k","f","h","ch")
complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")
simplexCs.ipa=c("m","n","l","w","j","w","ṭ","g","p","t","k","f","h","tʃ")
complexCs.ipa=c("ṭp","tp","kp","km","kṇ","mp","ɣ","lβʲ")

@@

#read.table("substitution_bank_try1.txt",header=T)->sub_bank
read.table("substitution_bank.txt",header=T)->sub_bank
colnames(sub_bank)<-c("target","realization")
sub_bank$target=factor(sub_bank$target,levels=allVC)
sub_bank$realization=factor(sub_bank$realization,levels=allVC)

@@

#read.table("deletion_bank_try1.txt",header=T)->del_bank
read.table("deletion_bank.txt",header=T)->del_bank
colnames(del_bank)<-c("target")
del_bank$target=factor(del_bank$target,levels=allVC)
del_tab= table(del_bank)[allVC]

#read.table("match_bank_try1.txt",header=T)->match_bank
read.table("match_bank.txt",header=T)->match_bank
colnames(match_bank)<-c("target")
match_bank$target=factor(match_bank$target,levels=allVC)
match_tab= table(match_bank)[allVC]

@@

Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data. Deletions were relatively rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets). Substitutions were somewhat more frequent, but still relatively rare. Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets)). Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (`r sum(sub_tab[oralVs,oralVs])` oral vowels (~`r mean(sub_pc[oralVs,oralVs],na.rm=T)`% of targets and `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels (~`r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% of targets). Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets. Interestingly, simple consonants were sometimes also substituted by complex ones; this happened `r sum(sub_tab[simplexCs,complexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all simple consonant targets. Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets). Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.
